Widespread nitrous oxide undersaturation in U.S. lakes and reservoirs

### 2023-12-14 manuscript/manuscript_file.Rmd
Updated manuscript file and support file with revised results

```{r source_scripts, echo=FALSE, warning=FALSE, message=FALSE}
# source scripts------
# some libraries are called within this script.  Not sure if that is necessary
# if libraries are also called above.  Thus far I'm including necessary
# library calls in external script.
# see above for .Rmd working directory
source("manuscript/manuscript_support.R", local = knitr::knit_global())

@@

**Lakes, ponds, and reservoirs are estimated to be globally important sources of nitrous oxide (N~2~O) to the atmosphere [@DelSontro_etal_2018a], but recent evidence of N~2~O uptake across a broad range of waterbodies have called the accuracy of emission estimates into question. Here we use a new national-scale dataset on dissolved N~2~O and a Bayesian hierarchical model to predict N~2~O concentration and emission rates in `r all_predictions %>% distinct(.row) %>% summarize(nlake = n()) %>% pull() %>% format(., big.mark=",", scientific=FALSE)` waterbodies in the conterminous U.S. (CONUS). We found that N~2~O undersaturation was widespread through the CONUS, with an estimated `r propN2oSinkMean`% (95% credible interval: `r propN2oSinkLower` - `r propN2oSinkUpper`%) of waterbodies functioning as N~2~O sinks. The model predicts dissolved N~2~O concentrations reasonably well based partly on interactions between nitrate (NO~3~) concentration, waterbody size, and water temperature. Despite working with the largest aquatic N~2~O dataset to date, our national-scale estimate of N~2~O emissions from CONUS waterbodies is poorly constrained, with a 95% credible interval ranging from net uptake to net emission (`r national_stats %>% select(LCL_fn2o) %>% pull() %>% round(., 0)` - `r national_stats %>% select(UCL_fn2o) %>% pull() %>% round(., 0)` metric tons N~2~O year^-1^). Widespread N~2~O undersaturation in CONUS waterbodies and a national-scale emission estimate that is too uncertain to determine if CONUS waterbodies are a net source or sink of N~2~O highlight the need to revisit N~2~O models which presume surface waters are a N~2~O source.**

@@

In this study we present the first national-scale survey of N~2~O concentrations and emissions from CONUS waterbodies. A generalized random-tessellation stratified survey design [@Stevens_Olsen_2004] was used to select a spatially-balanced and representative sample of `r dg %>% filter(sitetype == "PROB") %>% distinct(site.id) %>% summarize(n = n()) %>% pull()` sites from CONUS waterbodies that were at least 1 hectare in area, at least 1 meter deep, had at least 0.1 hectare of open water, and had a minimum residence time of one week[@EPAdesign] **(Fig. 1)**. Sample sites were distributed across all nine major CONUS ecoregions and varied widely in water chemistry, morphometry, watershed land-use, and climate [@epa_data]. `r undersatN2oPercent` percent of the sampled waterbodies were undersaturated in N~2~O and were therefore functioning as N~2~O sinks **(Fig. 1)**. These data, along with the survey design information, were used to develop a Bayesian hierarchical model to predict N~2~O concentrations for all `r all_predictions %>% distinct(.row) %>% summarize(nlake = n()) %>% pull() %>% format(., big.mark=",", scientific=FALSE)` CONUS waterbodies in the target population. At the population level, the majority of CONUS waterbodies (mean = `r propN2oSinkMean`%, 95% CI: `r propN2oSinkLower` - `r propN2oSinkUpper`) were predicted to be a net sink for N~2~O. The Western Mountains ecoregion had the smallest proportion of waterbodies functioning as N~2~O sinks (mean = `r propN2oSinkMeanWM`%, 95% CI: `r propN2oSinkLowerWM` - `r propN2oSinkUpperWM`) and the Northern Plains had the greatest (mean = `r propN2oSinkMeanNPL`%, 95% CI: `r propN2oSinkLowerNPL` - `r propN2oSinkUpperNPL`), although the 95% credible intervals overlap across all ecoregions **(Table 1)**.

@@

```{r table1, echo = FALSE}
# Set up data
table1_dat <- rbind(national_stats %>% 
        select(WSA9_NAME, 
               total_sa,
               estimate, LCL, UCL, 
               post_med_sat, LCI_sat, UCI_sat, 
               post_med_en2o, LCL_en2o, UCL_en2o, 
               post_med_fn2o, LCL_fn2o, UCL_fn2o),
      ecoreg_stats %>% 
        select(WSA9_NAME, 
               total_sa,
               estimate, LCL, UCL, 
               post_med_sat, LCI_sat, UCI_sat, 
               post_med_en2o, LCL_en2o, UCL_en2o,
               post_med_fn2o, LCL_fn2o, UCL_fn2o)) %>%
  mutate('N2O undersaturation extent' = paste0(estimate*100, " (", LCL*100, " - ", UCL*100, ")"),
         'N2O saturation ratio' = paste0(post_med_sat, " (", LCI_sat, " - ", UCI_sat, ")"),
         'N2O emission rate' = paste0(round(post_med_en2o, digits = 3), " (", round(LCL_en2o, digits = 3), " - ", round(UCL_en2o, digits = 3), ")"),
         'N2O flux' = paste0(round(post_med_fn2o, digits = 0), " (", round(LCL_fn2o, digits = 0), " - ", round(UCL_fn2o, digits = 0), ")")) %>% 
  select(WSA9_NAME, total_sa, 'N2O undersaturation extent', 'N2O saturation ratio', 'N2O emission rate', 'N2O flux') %>%
  rename('spatial domain' = WSA9_NAME,
         'waterbody surface area' = total_sa)

# create flextable
flextable(table1_dat) %>%
  set_header_labels('waterbody surface area' = "total waterbody surface area\n(hectares)",
                    'N2O undersaturation extent' = "N2O undersaturation extent\n(% waterbodies)",
                    'N2O emission rate' = "N2O emission rate\n(mg N2O m-2 d-1)",
                    'N2O flux' = "N2O flux\n(metric ton N2O y-1)") %>% # 1 megagram (Mg) = 1 1000 kg = 1 metric ton
  align(j=c(2:6), align = "center", part = "all") %>% # center align columns 2:4
  valign(valign = "top", part = "header") %>% # top align column names
  # caption doesn't show in .Rmd viewer, but renders correctly in .docx
  set_caption(
    caption = as_paragraph( # allows for formatting
      as_chunk("Table 1. Mean and 95% credible interval of N2O undersaturation extent, saturation ratio, emission rate, and flux at the national and ecoregional scales.", 
               props = fp_text_default(bold = TRUE) # here is where we specify formatting (font, size, bold, etc)
      )
    ),
    align_with_table = FALSE) %>% # table is center aligned, ignore that and apply default left align to caption

  footnote(i = 1, j = 3:6,
           ref_symbols = "a",
           value = as_paragraph("Mean and 95% credible interval of 2000 realizations of posterior predictive distribution"),
           part = "header"
  ) %>%
    footnote(i = 1, j = 3,
           ref_symbols = ", b",
           value = as_paragraph("Percent of population with N2O saturation ratio < 1."),
           part = "header",
           inline = TRUE
  ) %>%
  footnote(i = 1, j = 5:6,
           ref_symbols = ", c",
           value = as_paragraph("Negative values indicate N2O uptake"),
           part = "header",
           inline = TRUE
  ) %>%
  fontsize(size = 8, part = "all") %>% # doesn't apply to caption when as_paragraph is invoked.
  fontsize(size = 7, part = "footer") %>% #smaller font for footnote
    autofit() %>% # convenient column spacing
  fit_to_width(ft_1, max_width = 11) # scale to fill 8x11 paper, landscape

@@

Despite widespread N~2~O undersaturation in CONUS waterbodies, the mean predicted N~2~O saturation ratio (ratio of measured to equilibrium concentration) was statistically indistinguishable from one in all ecoregions **(Table 1, Fig. 2)**. The apparent contradiction between widespread N~2~O undersaturation and average dissolved N~2~O concentrations close to equilibrium can be attributed to the predicted distribution of dissolved N~2~O concentrations across the population of CONUS waterbodies **(Fig. 2)**. Across the CONUS, and within each of the nine major ecoregions, the predicted distribution of N~2~O saturation ratios had median values less than 1, indicating undersaturation, but a long right tail composed of relatively rare waterbodies with levels of N~2~O supersaturation that are sufficiently high to drive the population mean to values close to, or even greater than 1. Thus most waterbodies in the CONUS are functioning as N~2~O sinks during the summer months, but a few function as strong N~2~O sources, particularly at high NO~3~ at high water temperature **(Fig. 3)**.

![Figure 2. Posterior predictive distribution of N~2~O saturation values for waterbodies in the nine major ecoregions. Each blue line (n=2000) represents the distribution of values from a single realization of the posterior predictive distribution. Vertical black dashed and solid lines indicate point estimates of the median and mean, respectively (Table 1). Vertical red line indicates a saturation ratio of 1. x-axis is truncated at 3 to better show the area where 97% of the observations occur.](manuscript_figures/figure2.tiff)

@@

The model results suggest that NO~3~ may exert a strong control on N~2~O, but the relationship is complicated by interactions with water temperature and lake surface area **(Fig. 3)**. This finding mirrors reports from a survey of farm ponds in Canada where N~2~O was found to be associated with dissolved inorganic nitrogen (DIN) concentration, but the effect of DIN was conditional upon thermal stratification strength[@Webb_etal_2019]. These complex interactions may explain contrasting literature reports regarding the role of NO~3~ in regulating N~2~O. Several studies have reported positive correlations between NO~3~ and N~2~O in lakes [@Kortelainen_etal_2020; @Yang_etal_2015] while others have found no relationship [@Soued_etal_2016; @Whitfield_etal_2011]. It is possible that the NO~3~ \~ N~2~O relationship wasn't detected in the latter studies because it was confounded by interactions with temperature, lake size, thermal stratification, or other environmental variables.

The influence of NO~3~ levels on N~2~O saturation **(Fig. 3)** may reflect the relationship between NO~3~ levels and denitrification, a type of anaerobic metabolism that couples the oxidation of organic matter with the sequential reduction of NO~3~ to nitrite (NO~2~), nitric oxide (NO), N~2~O, and N~2~ [@Kuypers_etal_2018; @Richardson_etal_2009]. The reduction of NO~3~ to N~2~ can be incomplete, however, resulting in the release of N~2~O to the water column. Incomplete denitrification often occurs when NO~3~ is relatively abundant [@Beaulieu_etal_2015; @Beaulieu_etal_2014], likely because the final reduction of N~2~O to N~2~ is the least energetic reaction in the denitrification pathway, yielding less than 20% of the total energy yield of complete denitrification [@Koike_etal_1975; @Richardson_etal_2009]. When NO~3~ is limiting, however, denitrification can resort to scavenging N~2~O from the water column and reducing it to N~2~, thereby functioning as an N~2~O sink. N~2~O consumption via denitrification may explain the N~2~O undersaturation predicted at low NO~3~ levels, whereas denitrification may be a source of N~2~O at higher NO~3~ levels.

It is also possible that the correlation between N~2~O and NO~3~ reflects the simultaneous production of these solutes via nitrification. Nitrification, an aerobic process which can produce N~2~O as a byproduct, is the microbial oxidation of ammonium (NH~4~^+^) to NO~3~ [@Kuypers_etal_2018]. Nitrification may be more prevalent than denitrification in the predominantly well oxygenated surface waters where the N~2~O samples were collected because denitrification is an anaerobic process. However, denitrification can occur in anoxic microsites within particulate matter suspended in an aerobic water column [@Broman_etal_2021] and can be transported from aoxic sediments to the water surface via advection and diffusion [@DelSontro_etal_2018]. Thus the N~2~O - NO~3~ relationship may be reflective of some combination of N~2~O production/consumption via denitrification and N~2~O production via nitrification.

@@

The estimated pattern of increasing N~2~O saturation with water temperature **(Fig. 3)** is  consistent with nitrogen transformations occurring through temperature sensitive biological processes, but the relationship was dependent on NO~3~ levels. At low NO~3~ levels, temperature was estimated to have a relatively minor effect on N~2~O saturation, possibly because N~2~O production/consumption was limited by substrate rather than temperature. The effect of temperature was more pronounced at greater NO~3~ levels, potentially reflecting the stimulatory effect of higher temperature on biological N~2~O production. It appears that elevated water temperature may serve to magnify rates of N~2~O production that are largely dictated by NO~3~ availability.

Nitrous oxide saturation ratio was also related to waterbody size **(Fig. 3)**.  Small waterbodies with low NO~3~ levels were predicted to be undersaturated, whereas small waterbodies with high NO~3~ were predicted to be highly supersaturated. Large waterbodies, regardless of NO~3~ status, were predicted to have N~2~O saturation ratios closer to equilibrium. When pooled across all ecoregions, the absolute magnitude of the difference between the measured and equilibrium N~2~O concentration, known as $\Delta$ N~2~O, decreased with increasing waterbody size class **[Fig. 4]**. This pattern holds regardless of whether N~2~O was supersaturated or undersatured. Waterbody size and shape can affect dissolved gas concentration through several mechanisms. In waterbodies with low water volume to benthic surface area ratios, the mass of N~2~O consumed or produced in the sediments can be a substantial fraction of the mass of N~2~O dissolved in the water column [@Carpenter_1983]. Furthermore, sediment may warm faster in small and shallow waterbodies, whereas thermal stratification may substantially delay sediment warming in large and deep waterbodies, resulting in greater rates of N~2~O production or consumption in small lakes. Thermal stratification may also prevent the exchange of water between the lake surface and the hypolimnion, thereby preventing N~2~O depleted or enriched water from advecting to the lake surface [@Beaulieu_etal_2015; @Beaulieu_etal_2014].

@@

![Figure 4. Relationship between $\Delta$ N~2~O (difference between predicted dissolved and equilibrium N~2~O concentration) and waterbody surface areas. Mean (points) and 95% credible interview (horizontal bars) are derived from 2000 realizations of the posterior predictive distribution.](manuscript_figures/n2oStarBySize.tiff)

@@

The emission rates estimated across CONUS waterbodies span a broad range (`r min.max$min %>% round(.,2)` - `r min.max$max %>% round(.,2)` mg N~2~O m^-2^ d^-1^), which was expected for this large and diverse collection of waterbodies, but are within the range of values reported in the literature for lentic ecosystems [@Yang_etal_2015; @Kortelainen_etal_2020; @Webb_etal_2019]. Variability and magnitude of predicted N~2~O emission rates decreased with increasing waterbody size **(Fig. 5A)**, which is consistent with the relationship between N~2~O saturation and waterbody size **(Fig. 3C,D)**. Despite predicted widespread N~2~O undersaturation in CONUS waterbodies, the estimated national average N~2~O emission rate was positive (mean = `r national_stats$post_med_en2o`, 95% CI: `r national_stats$LCL_en2o` - `r national_stats$UCL_en2o` mg N~2~O m^-2^ d^-1^) **(Table 1)**. This contradiction is due to relatively rare waterbodies with very high emission rates that pull the mean (`r national_stats$post_med_en2o` mg N~2~O m^-2^ d^-1^) to a value that far exceeded the median (`r national_stats$post_median_en2o` 95% CI: `r national_stats$LCL_median_en2o` - `r national_stats$UCL_median_en2o` mg N~2~O m^-2^ d^-1^) **(SI Fig. 1)**.

![Figure 5: A) N~2~O emission rate and B) flux plotted against waterbody size for all 465,896 waterbodies in the target population. Across all plots, points and bars reflect the mean (points) and 95% credible interval, respectively, of the 2000 realizations of the posterior predictive distribution.](manuscript_figures/n2oFluxAndEmissionRateVsContinuousArea.tiff)

@@

Our research is the first to suggest that most lakes and reservoirs in the CONUS function as N~2~O sinks during the summer.  Our observations of dissolved N~2~O were predicted reasonably well using non-linear interactions between observed water temperature, NO~3~ concentration, and waterbody size (see SI). The model form and estimated parameters suggest that NO~3~ concentration plays a role in determining whether a waterbody functions as a sink or source of N~2~O, and the magnitude of the source or sink is exaggerated with increasing water temperature. The degree of N~2~O disequilibrium also increases with decreasing waterbody size, possibly because smaller waterbodies have low air-water gas exchange rates.

>>>>

```{r source_scripts, message=FALSE, warning=FALSE, include=FALSE}
# Load results objects from disk for knitting (if they already exist) ----------
# vector of object names
objs <- c("dg",
          "national_stats",
          "ecoreg_stats",
          "em_min_max",
          "FluxBySize",
          "ipcc.n2o"
          )

N <- length(objs)

# list of file paths for objects saved
f <- list()
for(i in 1:N){
    f[i] <- paste0(Sys.getenv("RENV_PROJECT"), "/manuscript/manuscript_files/",
    objs[i], ".rda")
    }

# load files if they exist
for(i in 1:N){
  if(file.exists(f[[i]]))
    load(f[[i]])
  else
  # source scripts------
  # some libraries are called within this script.  Not sure if that is necessary
  # if libraries are also called above.  Thus far I'm including necessary
  # library calls in external script.
  # see above for .Rmd working directory
    source("manuscript/manuscript_support.R", local = knitr::knit_global())
}

# clean up workspace
rm(f, i, objs, N)

# load sample frame from model files folder
load("modelFiles/sframe.rda")

# remove the Great Salt Lake from sample frame
sframe <- sframe[-258807,]

# Check for figures. If not available, load and run source script --------

# vector of figure names
objs <- c("cond_effect_no3.tif",
          "figure1.png",
          "figure2.tiff",
          "n2oFluxAndEmissionRateVsContinuousArea.tiff",
          "n2oStarBySize.tiff",
          "SIfigure1.tiff"
          )

N <- length(objs)

# list of file paths for objects saved
f <- list()
for(i in 1:N){
    f[i] <- paste0(Sys.getenv("RENV_PROJECT"), "/manuscript/manuscript_figures/",
    objs[i])
    }

# load files if they exist
for(i in 1:N){
  if(file.exists(f[[i]]))
    print("Good! Figure already created")
  else
    print(paste0("missing figure: ", f[[1]]))
 }

# clean up workspace
rm(f, i, objs, N)
```

@@

**Lakes, ponds, and reservoirs are estimated to be globally important sources of nitrous oxide (N~2~O) to the atmosphere [@DelSontro_etal_2018a], but recent evidence of N~2~O uptake across a broad range of waterbodies have called the accuracy of emission estimates into question. Here we use a new national-scale dataset on dissolved N~2~O and a Bayesian hierarchical model to predict N~2~O concentration and emission rates in `r sframe %>% summarize(nlake = n()) %>% pull() %>% format(., big.mark=",", scientific=FALSE)` waterbodies in the conterminous U.S. (CONUS). We found that N~2~O undersaturation was widespread through the CONUS, with an estimated `r national_stats$post_m_prop_sat * 100`% (95% credible interval: `r national_stats$LCL_prop_sat * 100` - `r national_stats$UCL_prop_sat * 100`%) of waterbodies functioning as N~2~O sinks. The model predicts dissolved N~2~O concentrations reasonably well based partly on interactions between nitrate (NO~3~) concentration, waterbody size, and water temperature. Despite working with the largest aquatic N~2~O dataset to date, our national-scale estimate of N~2~O emissions from CONUS waterbodies is poorly constrained, with a 95% credible interval ranging from net uptake to net emission (`r national_stats %>% select(LCL_fn2o) %>% pull() %>% round(., 0)` - `r national_stats %>% select(UCL_fn2o) %>% pull() %>% round(., 0)` metric tons N~2~O year^-1^). Widespread N~2~O undersaturation in CONUS waterbodies and a national-scale emission estimate that is too uncertain to determine if CONUS waterbodies are a net source or sink of N~2~O highlight the need to revisit N~2~O models which presume surface waters are a N~2~O source.**

@@

In this study we present the first national-scale survey of N~2~O concentrations and emissions from CONUS waterbodies. A generalized random-tessellation stratified survey design [@Stevens_Olsen_2004] was used to select a spatially-balanced and representative sample of `r dg %>% filter(sitetype == "PROB") %>% distinct(site.id) %>% summarize(n = n()) %>% pull()` sites from CONUS waterbodies that were at least 1 hectare in area, at least 1 meter deep, had at least 0.1 hectare of open water, and had a minimum residence time of one week[@EPAdesign] **(Fig. 1)**. Sample sites were distributed across all nine major CONUS ecoregions and varied widely in water chemistry, morphometry, watershed land-use, and climate [@epa_data]. `r round(((dg %>% dplyr::filter(n2o.src.snk == "sink") %>% {nrow(.)} / dg %>% distinct(site.id) %>% {nrow(.)}) * 100), 1)` percent of the sampled waterbodies were undersaturated in N~2~O and were therefore functioning as N~2~O sinks **(Fig. 1)**. These data, along with the survey design information, were used to develop a Bayesian hierarchical model to predict N~2~O concentrations for all `r sframe %>% summarize(nlake = n()) %>% pull() %>% format(., big.mark=",", scientific=FALSE)` CONUS waterbodies in the target population. At the population level, the majority of CONUS waterbodies (posterior mean = `r national_stats$post_m_prop_sat * 100`%, 95% CI: `r national_stats$LCL_prop_sat * 100` - `r national_stats$UCL_prop_sat * 100`) were predicted to be a net sink for N~2~O. The Western Mountains ecoregion had the smallest proportion of waterbodies predicted as N~2~O sinks (posterior mean = `r ecoreg_stats %>% filter(WSA9 == "WMT") %>% pull(post_m_prop_sat) * 100`%, 95% CI: `r ecoreg_stats %>% filter(WSA9 == "WMT") %>% pull(LCL_prop_sat) * 100` - `r ecoreg_stats %>% filter(WSA9 == "WMT") %>% pull(UCL_prop_sat) * 100`) and the Northern Plains had the greatest (posterior mean = `r ecoreg_stats %>% filter(WSA9 == "NPL") %>% pull(post_m_prop_sat) * 100`%, 95% CI: `r ecoreg_stats %>% filter(WSA9 == "NPL") %>% pull(LCL_prop_sat) * 100` - `r ecoreg_stats %>% filter(WSA9 == "NPL") %>% pull(UCL_prop_sat) * 100`), although the 95% credible intervals over the predictive distributions overlap across all ecoregions **(Table 1)**.

@@

```{r table1, echo = FALSE}
# Set up data
table1_dat <- rbind(national_stats %>% 
        select(WSA9_NAME, 
               total_sa,
               post_m_mean_sat, LCL_mean_sat, UCL_mean_sat, 
               post_m_prop_sat, LCL_prop_sat, UCL_prop_sat, 
               post_m_mean_en2o, LCL_mean_en2o, UCL_mean_en2o, 
               post_m_fn2o, LCL_fn2o, UCL_fn2o),
      ecoreg_stats %>% 
        select(WSA9_NAME, 
               total_sa,
               post_m_mean_sat, LCL_mean_sat, UCL_mean_sat, 
               post_m_prop_sat, LCL_prop_sat, UCL_prop_sat, 
               post_m_mean_en2o, LCL_mean_en2o, UCL_mean_en2o,
               post_m_fn2o, LCL_fn2o, UCL_fn2o)) %>%
  mutate('Mean N2O saturation ratio' = paste0(post_m_mean_sat,
                                         " (", 
                                         LCL_mean_sat,
                                         " - ", 
                                         UCL_mean_sat, ")"),
         'N2O undersaturation extent' = paste0(post_m_prop_sat * 100,
                                               " (", 
                                               LCL_prop_sat * 100, 
                                               " - ", 
                                               UCL_prop_sat * 100,
                                               ")"),
         'N2O emission rate' = paste0(round(post_m_mean_en2o, digits = 3),
                                      " (", 
                                      round(LCL_mean_en2o, digits = 3), 
                                      " - ", 
                                      round(UCL_mean_en2o, digits = 3), 
                                      ")"),
         'N2O flux' = paste0(round(post_m_fn2o, digits = 0), 
                             " (", 
                             round(LCL_fn2o, digits = 0), 
                             " - ",
                             round(UCL_fn2o, digits = 0),
                             ")")) %>% 
  select(WSA9_NAME, total_sa, 'Mean N2O saturation ratio', 'N2O undersaturation extent', 'N2O emission rate', 'N2O flux') %>%
  rename('spatial domain' = WSA9_NAME,
         'waterbody surface area' = total_sa)

# create flextable
flextable(table1_dat) %>%
  set_header_labels('waterbody surface area' = "total waterbody surface area\n(hectares)",
                    'N2O undersaturation extent' = "N2O undersaturation extent\n(% waterbodies)",
                    'N2O emission rate' = "N2O emission rate\n(mg N2O m-2 d-1)",
                    'N2O flux' = "N2O flux\n(metric ton N2O y-1)") %>% # 1 megagram (Mg) = 1 1000 kg = 1 metric ton
  align(j=c(2:6), align = "center", part = "all") %>% # center align columns 2:4
  valign(valign = "top", part = "header") %>% # top align column names
  # caption doesn't show in .Rmd viewer, but renders correctly in .docx
  set_caption(
    caption = as_paragraph( # allows for formatting
      as_chunk("Table 1. Mean saturation ratio, undersaturation extent, emission rate, and flux estimates for N2O at the national and ecoregional scales.", 
               props = fp_text_default(bold = TRUE) # here is where we specify formatting (font, size, bold, etc)
      )
    ),
    align_with_table = FALSE) %>% # table is center aligned, ignore that and apply default left align to caption

  footnote(i = 1, j = 3:6,
           ref_symbols = "a",
           value = as_paragraph("Mean and 95% credible intervals based on 2000 draws from the posterior distribution"),
           part = "header"
  ) %>%
    footnote(i = 1, j = 4,
           ref_symbols = ", b",
           value = as_paragraph("Percent of population with N2O saturation ratio < 1."),
           part = "header",
           inline = TRUE
  ) %>%
  footnote(i = 1, j = 5:6,
           ref_symbols = ", c",
           value = as_paragraph("Negative values indicate N2O uptake"),
           part = "header",
           inline = TRUE
  ) %>%
  fontsize(size = 8, part = "all") %>% # doesn't apply to caption when as_paragraph is invoked.
  fontsize(size = 7, part = "footer") %>% #smaller font for footnote
    autofit() %>% # convenient column spacing
  fit_to_width(ft_1, max_width = 11) # scale to fill 8x11 paper, landscape
```

@@

Despite widespread N~2~O undersaturation across CONUS waterbodies, the estimated means for N~2~O saturation ratio (ratio of measured to equilibrium concentration) were indistinguishable from one in all ecoregions **(Table 1, Fig. 2)**. The apparent contradiction of widespread N~2~O undersaturation with average dissolved N~2~O concentrations closer to equilibrium can be attributed to the prominent right skew predicted for dissolved N~2~O concentrations across the population of CONUS waterbodies **(Fig. 2)**. Across the CONUS, and within each of the nine major ecoregions, each predicted distribution of N~2~O saturation ratios had a median value less than 1, indicating that most waterbodies were undersaturated in these populations. However, the long right tails on these distributions, comprised of relatively rare waterbodies with very high levels of N~2~O supersaturation, were sufficient to drive the population means to values close to, or even greater than 1. Thus, most waterbodies in the CONUS were functioning as N~2~O sinks during the summer months, but a few functioned as strong N~2~O sources, particularly at high NO~3~ and high water temperature **(Fig. 3)**.

![Figure 2. Predicted distribution of N~2~O saturation values across waterbodies in the nine major ecoregions. Each blue line (n=2000) represents the distribution of values from a single realization of the posterior predictive distribution. Vertical black dashed and solid lines indicate point estimates of the median and mean, respectively, across all realizations (Table 1). Vertical red line indicates a saturation ratio of 1. x-axis is truncated at 3 to better show the area where 97% of the observations occurred.](manuscript_figures/figure2.tiff)

@@

The model results suggest that NO~3~ may be an important predictor for N~2~O, but the relationship is complicated by interactions with water temperature and lake surface area **(Fig. 3)**. This finding mirrors reports from a survey of farm ponds in Canada where N~2~O was found to be associated with dissolved inorganic nitrogen (DIN) concentration, but the effect of DIN was conditional upon thermal stratification strength[@Webb_etal_2019]. These complex interactions may explain contrasting literature reports regarding the role of NO~3~ in regulating N~2~O. Several studies have reported positive correlations between NO~3~ and N~2~O in lakes [@Kortelainen_etal_2020; @Yang_etal_2015] whereas others have found no relationship [@Soued_etal_2016; @Whitfield_etal_2011]. It is possible that the NO~3~ \~ N~2~O relationship wasn't detected in the latter studies because it was confounded by interactions with temperature, lake size, thermal stratification, or other environmental variables.

The influence of NO~3~ levels on N~2~O saturation **(Fig. 3)** may reflect a relationship between NO~3~ levels and denitrification, a type of anaerobic metabolism that couples the oxidation of organic matter with the sequential reduction of NO~3~ to nitrite (NO~2~), nitric oxide (NO), N~2~O, and N~2~ [@Kuypers_etal_2018; @Richardson_etal_2009]. The reduction of NO~3~ to N~2~ can be incomplete, however, resulting in the release of N~2~O to the water column. Incomplete denitrification often occurs when NO~3~ is relatively abundant [@Beaulieu_etal_2015; @Beaulieu_etal_2014], likely because the final reduction of N~2~O to N~2~ is the least energetic reaction in the denitrification pathway, yielding less than 20% of the total energy yield of complete denitrification [@Koike_etal_1975; @Richardson_etal_2009]. When NO~3~ is limiting, however, denitrification can resort to scavenging N~2~O from the water column and reducing it to N~2~, thereby functioning as an N~2~O sink. N~2~O consumption via denitrification may explain the N~2~O undersaturation predicted at low NO~3~ levels, whereas denitrification may be a source of N~2~O at higher NO~3~ levels.

It is also possible that the correlation between N~2~O and NO~3~ reflects the simultaneous production of these solutes via nitrification. Nitrification, an aerobic process which can produce N~2~O as a byproduct, is the microbial oxidation of ammonium (NH~4~^+^) to NO~3~ [@Kuypers_etal_2018]. Nitrification may be more prevalent than denitrification in the predominantly well oxygenated surface waters where the N~2~O samples were collected because denitrification is an anaerobic process. However, denitrification can occur in anoxic microsites within particulate matter suspended in an aerobic water column [@Broman_etal_2021] and can be transported from anoxic sediments to the water surface via advection and diffusion [@DelSontro_etal_2018]. Thus the N~2~O - NO~3~ relationship may be reflective of some combination of N~2~O production/consumption via denitrification and N~2~O production via nitrification.

@@

The estimated pattern of increasing N~2~O saturation with water temperature **(Fig. 3)** is  consistent with nitrogen transformations occurring through temperature sensitive biological processes, but the relationship was dependent on NO~3~ levels. At low NO~3~ levels, temperature was estimated to have a relatively minor effect on N~2~O saturation, possibly because N~2~O production/consumption was limited by substrate rather than temperature. The effect of temperature was more pronounced at greater NO~3~ levels, potentially reflecting the stimulatory effect of higher temperature on biological N~2~O production. Elevated water temperature may serve to magnify rates of N~2~O production that may be largely dictated by NO~3~ availability.

Nitrous oxide saturation ratio was also related to waterbody size **(Fig. 3)**.  Small waterbodies with low NO~3~ levels were predicted to be undersaturated, whereas small waterbodies with high NO~3~ were predicted to be highly supersaturated. Large waterbodies, regardless of NO~3~ status, were predicted to have N~2~O saturation ratios closer to equilibrium. When pooled across all ecoregions, the absolute magnitude of the difference between the measured and equilibrium N~2~O concentration, known as $\Delta$ N~2~O, decreased with increasing waterbody size class **(Fig. 4)**. This pattern holds regardless of whether N~2~O was supersaturated or undersaturated. Waterbody size and shape can affect dissolved gas concentration through several mechanisms. In waterbodies with low water volume to benthic surface area ratios, the mass of N~2~O consumed or produced in the sediments can be a substantial fraction of the mass of N~2~O dissolved in the water column [@Carpenter_1983]. Furthermore, sediment may warm faster in small and shallow waterbodies, whereas thermal stratification may substantially delay sediment warming in large and deep waterbodies, resulting in greater rates of N~2~O production or consumption in small lakes. Thermal stratification may also prevent the exchange of water between the lake surface and the hypolimnion, thereby preventing N~2~O depleted or enriched water from advecting to the lake surface [@Beaulieu_etal_2015; @Beaulieu_etal_2014].

@@

![Figure 4. Estimated mean $\Delta$ N~2~O (difference between dissolved and equilibrium N~2~O concentrations) versus waterbody surface area (ha) size categories. Posterior mean (points) and 95% credible intervals (horizontal bars) were based on 2000 draws from the posterior distribution.](manuscript_figures/n2oStarBySize.tiff)

@@

The predicted emission rates for CONUS waterbodies spanned a broad range (95% CI min: `r em_min_max$min_LCL` - `r em_min_max$min_UCL` mg N~2~O m^-2^ d^-1^, 95% CI max: `r em_min_max$max_LCL` - `r em_min_max$max_UCL` mg N~2~O m^-2^ d^-1^), which was expected for this large and diverse collection of waterbodies, but within the range of values reported in the literature for lentic ecosystems [@Yang_etal_2015; @Kortelainen_etal_2020; @Webb_etal_2019]. Variability and magnitude of predicted N~2~O emission rates decreased with increasing waterbody size **(Fig. 5A)**, which reflects the relationship between N~2~O saturation and waterbody size **(Fig. 3C,D)**. Despite widespread N~2~O undersaturation predicted across CONUS waterbodies, the estimate for the national average N~2~O emission rate was centered over positive values (posterior mean = `r national_stats$post_m_mean_en2o`, 95% CI: `r national_stats$LCL_mean_en2o` - `r national_stats$UCL_mean_en2o` mg N~2~O m^-2^ d^-1^) **(Table 1)** and clearly exceeded the negative estimate for the national median (posterior mean = `r national_stats$post_m_median_en2o`, 95% CI: `r national_stats$LCL_median_en2o` - `r national_stats$UCL_median_en2o` mg N~2~O m^-2^ d^-1^). This departure from the median was due to relatively rare waterbodies with very high emission rates pulling the mean towards the direction of positive skew **(SI Fig. 1)**.

![Figure 5: A) Predicted N~2~O emission rate and B) flux plotted against waterbody size for all 465,896 waterbodies in the target population. Points and bars reflect the posterior means (points) and 95% credible intervals, respectively, based on 2000 draws from the posterior predictive distribution for each waterbody. C) A comparison of the estimated relative flux contributions from waterbodies in each size class to the total CONUS flux. The red line connects the mean of the posterior estimate of the ratio, at each size class, based on 2000 draws. The black polygon indicates the 95% credible intervals for the ratio estimates across the size classes](manuscript_figures/n2oFluxAndEmissionRateVsContinuousArea.tiff)

@@

Our research is the first to suggest that most lakes and reservoirs in the CONUS function as N~2~O sinks during the summer.  Our observations of dissolved N~2~O were predicted reasonably well using a regression model with interactions between observed water temperature, NO~3~ concentration, and waterbody size (see SI). These modeling results suggested that NO~3~ levels may play a role in determining whether a waterbody functions as a sink or source of N~2~O, and the magnitude of the source or sink may be exaggerated with increasing water temperature. The degree of N~2~O disequilibrium also increased with decreasing waterbody size, possibly because smaller waterbodies have low air-water gas exchange rates.

Although waterbodies in the largest lake size class tend to have N~2~O levels close to equilibrium, they emit more N~2~O than the other size classes due to their expansive surface area. The N~2~O flux from these waterbodies is also more uncertain than the smaller lake size classes because relatively small uncertainties in dissolved and equilibrium N~2~O concentrations translate to large uncertainties in flux when propagated across the surface of large waterbodies. Uncertainty in the national flux estimate was too large to determine if CONUS waterbodies functioned as a net source or sink of N~2~O. While our sample included some of the largest waterbodies in the CONUS, efforts to improve the accuracy of the lentic N~2~O flux estimate should include an assessment of intra-waterbody spatial variation in N~2~O, which can be important in large waterbodies[@Salk_etal_2019; @Borges_etal_2023], as well as careful accounting of measurement an analytical uncertainty, which can propagate to uncertainties in flux estimates.
